---
title: "Chapter4.Rmd"
author: "Kevin Sandeman"
date: "22 november 2017"
output: html_document
---
# RStudio Exercise 4

*This week clustering and classification are central in this course.*

##Create this file and describe the data
```{r}
library(MASS)
data("Boston")
str(Boston)
```
 In the "Boston"-data set there are 506 observations with the following 14 variables:  
 
1. **crim:** per capita crime rate by town.  
2. **zn:** proportion of residential land zoned for lots over 25,000 sq.ft.  
3. **indus:** proportion of non-retail business acres per town.  
4. **chas:** Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).  
5. **nox:** nitrogen oxides concentration (parts per 10 million).  
6. **rm:** average number of rooms per dwelling.  
7. **age:** proportion of owner-occupied units built prior to 1940.  
8. **dis:** weighted mean of distances to five Boston employment centers.  
9. **rad:** index of accessibility to radial highways.  
10. **tax:** full-value property-tax rate per \$10,000.  
11. **ptratio:** pupil-teacher ratio by town.  
12. **black:** 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town.  
13. **lstat:** lower status of the population (percent).  
14. **medv:** median value of owner-occupied homes in \$1000s.  

##Grafical overview and summary

```{r}
pairs(Boston)
```

This plot isn't very helpful for interpretation of the data, but shows some correlative relationships. A clearer statement about correlations will be made based upon the correlationsplot and -table below.

```{r}
summary(Boston)
```

In the summary there are som non-parametrically distributed parameters as crim, zn, rad, age, tax, indus and black. Other parameters can be distributed parameterically. In case of doubt a test for normality could be done.

```{r}
library(tidyverse)
library(corrplot)
cor_matrix<-cor(Boston) %>% round(digits = 2)
cor_matrix
corrplot(cor_matrix, method="circle", type = "upper", cl.pos= "b", tl.pos = "d", tl.cex = 0.6)
```

There seem to be negative correlations between dis versus age, nox, indus and tax, suggesting that the further from the employment centers the buildings are younger, there is less nitrous oxide pollution ,the proportion of non-retail businesses is lower and the full-value property-tax rate is higher. The correlation between medv and lstat is outspoken negative, meaning that areas with lower social status of the population have cheaper owner-occupied houses and vice versa.  

##Standarized data set and division into training and test set

```{r}
boston_scaled <- scale(Boston)
summary(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)
```

The data is scaled to the values above to eliminate the effect of different values on the following analyses. The data is divided into a training and test set.

```{r}

# create a quantile vector of crim
bins <- quantile(boston_scaled$crim)
# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins,  labels = c("low", "med_low", "med_high", "high"), include.lowest = TRUE)
# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)
# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)
# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)
# create train set
train <- boston_scaled[ind,]
# create test set 
test <- boston_scaled[-ind,]
```

##Linear discriminant analysis

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)
# print the lda.fit object
lda.fit
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)
# plot the lda results
plot(lda.fit , dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 2)
```
  
In this example the LD1-model explains 95 % of the clustering with the index of accessibility to radial highways (rad) as the most influential linear separator.

##Predict LDA model on test data

```{r}
# save the correct classes from test data
correct_classes <- test$crime
# remove the crime variable from test data
test <- dplyr::select(test, -crime)
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
```

The model predicts well the true values for the categorical crime rate. Most of the wrong predictions were adressed to the med_low crime group.

##k-means algoritm

```{r}
#reload Boston and scale
boston_scaled <- scale(Boston)
boston_scaled <- as.data.frame(boston_scaled)
summary(boston_scaled)

# scaled distance matrix
dist_scaled <- dist(boston_scaled)
# look at the summary of the distances
summary(dist_scaled)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(dist_scaled, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <-kmeans(dist_scaled, centers = 2)

# plot the Boston dataset with clusters

km <-kmeans(dist_scaled, centers = 2)
pairs(boston_scaled, col = km$cluster)
```

The optimal number for clusters, according to k-means algorithm is 2. In the correlation plot above, we can see that the plots with strong correlations between the parameters, show for most of the variables a division into two clusters, according to k-means method.

##K-means with 4 clusters and LDA

```{r}
km <-kmeans(dist_scaled, centers = 4)
lda.fit <- lda(km$cluster~., data = boston_scaled)
lda.fit
```

```{r}
plot(lda.fit, dimen = 2, col=classes, pch=classes)
lda.arrows(lda.fit, myscale = 3)
```


In this example the LD1-model explains only 76 % of the clustering with rad, nox, indus and zn as the most influential linear separators. K-mean algorithm with 4 centers isn't a better model to cluster this data.

##Plotly

```{r}
model_predictors <- dplyr::select(train, -crime)
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)
#k-means
dist_train <- dist(train)
km <-kmeans(dist_train, centers = 2)

# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)

#Matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
library("plotly")
#Create a 3D plot of the columns of the matrix product
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color=train$crime)
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color=km$cluster)
```

We can see that the high risk and med_high risk crime groups from the LDA-trained set overlap with one cluster according to the k-means algorithm for two cluster. The other cluster, according to k-means algoritm overlaps with low and med_low crime groups from the LDA-trained set.
